### 2016-07-05 abstract.tex
KC's comments.

  Serializability is a well-understood mechanism that simplifies
  reasoning about the behavior of concurrent transactions by ensuring
  transactions are \emph{isolated} from each other's effects once they
  begin executing.  However, enforcing serializability comes at a
  steep cost in performance because it necessarily restricts
  opportunities to exploit concurrency even when such opportunities
  would not violate application-specific invariants.

  As a result, database systems (in practice) support, and often
  encourage, developers to use weaker alternatives.  These
  alternatives break the strong isolation guarantees offered by
  serializable transactions to permit greater concurrency.
  Unfortunately, the semantics of weak isolation is poorly understood,
  and usually explained only informally in terms of low-level
  implementation artifacts.  Consequently, verifying high-level
  correctness properties in such environments is problematic.

  To address this issue, we present a program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions.  Notably, our development is
  parametric over a transaction's specific isolation semantics, and
  the consistency guarantees provided by the underlying data store,
  allowing it to be applicable over a large range of concurrency
  control mechanisms.  Case studies and experiments on real-world
  applications demonstrate the utility of our approach, and provide
  strong evidence that weakly-isolated transactions can be placed on
  the same formal footing as their strongly-isolated serializable
  counterparts.

>>>>

  Serializability is a well-understood correctness criterion that
  simplifies reasoning about the behavior of concurrent transactions
  by ensuring that transactions are \emph{isolated} from each other.
  However, enforcing serializable isolation comes at a steep cost in
  performance because it necessarily restricts opportunities to
  exploit concurrency even when such opportunities would not violate
  application-specific invariants. As a result, database systems in
  practice support, and often encourage, developers to use weaker
  alternatives. These alternatives break the strong isolation
  guarantees offered by serializable transactions to permit greater
  concurrency. Unfortunately, the semantics of weak isolation is
  poorly understood, and usually explained only informally in terms of
  low-level implementation artifacts. Consequently, verifying
  high-level correctness properties in such environments remains a
  challenging problem.

  To address this issue, we present a program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions. Notably, our development is parametric
  over a transaction's specific isolation semantics, and the
  consistency guarantees provided by the underlying data store,
  allowing it to be applicable over a range of concurrency control
  mechanisms.  Case studies and experiments on real-world applications
  demonstrate the utility of our approach, and provide strong evidence
  that weakly-isolated transactions can be placed on the same formal
  footing as their strongly-isolated serializable counterparts.

### 2016-07-05 introduction.tex
KC's comments.

Database transactions allow users to group operations on multiple
objects into a coherent unit, equipped with a set of four key
properties - atomicity, consistency, isolation, and durability (ACID).
Concurrency control mechanisms provide specific instantiations of
these properties to yield different ACID variants that characterize
how and when the effects of concurrently executing transactions become
visible to one another.  \emph{Serializability} is a particularly
well-studied instantiation that imposes strong atomicity and isolation
constraints on transaction execution, ensuring that any permissible
concurrent schedule yields results equivalent to a serial one in which
there is no interleaving of actions from different transactions.

The guarantees provided by serializability, and strong isolation in
particular, do not come for free, however - pessimistic concurrency
control methods that enforce serializability, for example, require
databases to provide expensive techniques such as two-phase locking
that incur overhead to deal with deadlocks, rollbacks, and
re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
optimistic multi-version concurrency control methods that must deal
with timestamp and version management~\cite{BG81}.  Regardless of the
particular concurrency control method employed, enforcing
serializability in a replicated data store additionally requires
global coordination among geo-distributed replicas, which makes the
store \emph{unavailable} in the presence of network
partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.

The tension between serializable transactions, which are easy to
reason about but difficult to implement, and more pragmatic variants
that are driven by performance and availability considerations, has
motivated the development of weaker forms of transaction isolation,
beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
defines three such weak isolation levels which are now implemented in
many relational and NoSQL databases. Not surprisingly, weakly-isolated
transactions have been found to significantly outperform serializable
transactions on benchmark suites, both on single-node databases and
multi-node replicated stores~\cite{dbtuningbook,bailishat,bailisvldb},
leading to their overwhelming adoption. A 2013
study~\cite{bailishotos} of 18 popular ACID and ``NewSQL'' databases
found that only three of them offer serializability by default, and
half, including Oracle 11g, do not offer it at all.  A 2015
study~\cite{bailisferal} of a large corpus of database applications
finds no evidence that applications manifestly change the default
isolation level offered by the database. Taken together, these studies
make clear that weakly-isolated transactions are quite prevalent in
practice, and strongly-isolated (serializable) transactions are often
eschewed altogether.

@@

Recent results on reasoning about application invariants in the
presence of weak \emph{consistency}~\cite{burckhardt14, redblueosdi,
  redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
At a high-level, reasoning about weak consistency involves reasoning
about the concurrent execution of strongly atomic and fully isolated
operations against different (potentially replicated), but consistent
states of the database or data store, where each operation locally
preserves application invariants. In contrast, reasoning about weak
isolation involves reasoning about the concurrent execution of
transactions, whose operations witness different states which,
although consistent individually, may not be obviously reconciled into
a consistent global view.  Intuitively, the latter problem reduces to
the former if all transactions contain a single
operation. Furthermore, weak isolation can exist independent of weak
consistency, as evident from the presence of weakly isolated
transactions on sequentially consistent relational databases; such
systems existed long before the advent of weakly consistent NoSQL
stores.  Thus, frameworks for reasoning about weak isolation will
necessarily have to generalize the reasoning frameworks developed for
weak consistency in new and important ways.

@@

In this paper, we propose a program logic for weakly-isolated
transactions that realizes this goal.  In particular, we develop a set
of syntax-directed compositional proof rules that allow developers to
build correctness proofs for transactional programs in the presence of
a weakly isolated concurrency control mechanism.  A key novelty of our
approach is that it is parametric over the \emph{isolation semantics}
of transactions in the program, as well as the \emph{consistency
  semantics} of the underlying store. In concrete terms, this means
that, unlike recent work focused on formalizing the semantics of weak
consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
does not assume a minimum or predefined set of consistency or
isolation levels. Instead, our operational semantics admits
declarative specifications of transaction isolation and store
consistency levels (\eg~\cite{pldi15,gotsmanconcur15}), and generates
executions that are guaranteed to satisfy these specifications. The
result is a flexible system model that is general enough to
incorporate the semantics of a range of isolation levels on a variety
of stores (\eg the sequentially consistent store of~\cite{adyaphd}, or
the causally consistent store of~\cite{gotsmanpopl16}, etc).  Our key
technical contribution is thus a set of proof rules that demonstrate
that a program with a given selection of isolation levels for its
transactions preserves its invariants when executed on a store
equipped with certain consistency guarantees (e.g., causal
visibility).

>>>>

Database transactions allow users to group operations on multiple
objects into a single logical unit, equipped with a set of four key
properties - atomicity, consistency, isolation, and durability (ACID).
Concurrency control mechanisms provide specific instantiations of
these properties to yield different ACID variants that characterize
how and when the effects of concurrently executing transactions become
visible to one another.  \emph{Serializability} is a particularly
well-studied instantiation that imposes strong atomicity and isolation
constraints on transaction execution, ensuring that any permissible
concurrent schedule yields results equivalent to a serial one in which
there is no interleaving of actions from different transactions.

The guarantees provided by serializability do not come for free,
however - pessimistic concurrency control methods that enforce
serializability, for example, require databases to provide expensive
techniques such as two-phase locking that incur overhead to deal with
deadlocks, rollbacks, and re-execution~\cite{twopl,ullmanbook}.
Similar criticisms apply to optimistic multi-version concurrency
control methods that must deal with timestamp and version
management~\cite{BG81}.  Regardless of the particular concurrency
control method employed, enforcing serializability in a replicated
data store additionally requires global coordination among
geo-distributed replicas, which makes the store \emph{unavailable} in
the presence of network
partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.

The tension between serializable transactions, which are easy to
reason about but difficult to implement, and more pragmatic variants
that are driven by performance and availability considerations, has
motivated the development of weaker forms of transaction isolation,
beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
defines three such weak isolation levels which are now implemented in
many relational and NoSQL databases. Not surprisingly, weakly-isolated
transactions have been found to significantly outperform serializable
transactions on benchmark suites, both on single-node databases and
multi-node replicated stores~\cite{dbtuningbook,bailishat,bailisvldb},
leading to their overwhelming adoption. A 2013
study~\cite{bailishotos} of 18 popular ACID and ``NewSQL'' databases
found that only three of them offer serializability by default, and
half, including Oracle 11g, do not offer it at all.  A 2015
study~\cite{bailisferal} of a large corpus of database applications
finds no evidence that applications manifestly change the default
isolation level offered by the database. Taken together, these studies
make clear that weakly-isolated transactions are quite prevalent in
practice, whereas serializable transactions are often eschewed.

@@

Recent results on reasoning about application invariants in the
presence of weak \emph{consistency}~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that usually manifests on replicated
data stores, where operations are concurrently executed against
different replicas, resulting in an order of execution inconsistent
with their serial order. The operations, nonetheless, are atomic and
fully isolated, and each operation is required to locally preserve
application invariants. In contrast, weak isolation is a property of
transactions, which are sets of atomic operations. Weak isolation
manifests when successive atomic operations in a transaction witness
different contemporary states (or different replicas) of the database
which, although consistent individually, may not be obviously
reconciled into a consistent global view. Furthermore, operations may
temporarily break invariants as long as transaction as a whole
preserves them.  Intuitively, the latter problem reduces to the former
if all transactions contain a single operation. Furthermore, weak
isolation can exist independent of weak consistency, as evident from
the presence of weakly isolated transactions on sequentially
consistent relational databases; such systems existed long before the
advent of weakly consistent NoSQL stores.  Thus, frameworks for
reasoning about weak isolation will necessarily have to generalize the
reasoning frameworks developed for weak consistency in new and
important ways.

@@

In this paper, we propose a program logic for weakly-isolated
transactions that realizes this goal.  In particular, we develop a set
of syntax-directed compositional proof rules that allow developers to
build correctness proofs for transactional programs in the presence of
a weakly isolated concurrency control mechanism.  A key novelty of our
approach is that it is parametric over the \emph{isolation semantics}
of transactions in the program, as well as the \emph{consistency
semantics} of the underlying store. In concrete terms, this means
that, unlike recent work focused on reasoning about programs under
weak consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system
model does not assume a minimum or predefined set of consistency or
isolation levels. Instead, our operational semantics admits
declarative specifications of transaction isolation and store
consistency, and generates executions that are guaranteed to satisfy
these specifications. The result is a flexible system model that is
general enough to incorporate the semantics of a range of isolation
levels on a variety of stores (\eg the sequentially consistent store
of~\cite{adyaphd}, or the causally consistent store
of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
a set of proof rules that demonstrate that a program with a given
selection of isolation levels for its transactions preserves its
invariants when executed on a store equipped with certain consistency
guarantees (\eg causal consistency).
